---
title: "BAYESIAN INFERENCE"
author: "NEEMA NDANU"
date: "2025-01-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  QUESTION 1

Marjorie is worried about whether it is safe to visit a vulnerable relative during a pandemic. She is 
considering whether to take an at-home test for the virus before visiting her relative. Assume the test 
has sensitivity 85% and specificity 92%. That is, the probability that the test will be positive is about 
85% if an individual is infected with the virus, and the probability that test will be negative is about 
92% if an individual is not infected. 

### a. Assume that about 2 in every 1,000 persons in the population is currently infected. What is the posterior probability that an individual with a positive test has the disease?

```{r}
# Define parameters
sensitivity <- 0.85            
specificity <- 0.92            
prevalence_1 <- 2 / 1000         

# Compute probabilities
P_D1_given_T1 <- sensitivity     
P_D0_given_T1 <- 1 - specificity 
P_D0_given_T0 <- specificity     
P_D1_given_T0 <- 1 - sensitivity 
P_D1 <- prevalence_1             
P_D0 <- 1 - prevalence_1         

# Total probability of a positive test
P_T1 <- (P_D1_given_T1 * P_D1) + (P_D0_given_T1 * P_D0)  

# Posterior probability: P(D = 1 | T = 1)
P_infected_given_positive <- (P_D1_given_T1 * P_D1) / P_T1

# Print the result
cat("The posterior probability that an individual with a positive test has the disease (prevalence = 2 in 1,000) is:", 
    round(P_infected_given_positive, 4), "\n")

```
This value reflects the likelihood of being infected given a positive test result. The low prevalence contributed largely to the low posterior probability.

### Suppose case counts have decreased substantially to about 15 in 100,000. What is the posterior probability that an individual with a positive test has the disease? 

```{r}
# Define parameters
prevalence_2 <- 15 / 100000      
P_D1_1 <- prevalence_2           
P_D0_1 <- 1 - prevalence_2       

# Total probability of a positive test
P_T1_1 <- (P_D1_given_T1 * P_D1_1) + (P_D0_given_T1 * P_D0_1)

# Posterior probability: P(D = 1 | T = 1)
P_infected_given_positive_1 <- (P_D1_given_T1 * P_D1_1) / P_T1_1

# Print the result
cat("The posterior probability that an individual with a positive test has the disease (prevalence = 15 in 100,000) is:", 
    round(P_infected_given_positive_1, 4), "\n")

```
This value reflects the likelihood of being infected given a positive test result. The low prevalence contributed largely to the low posterior probability.

As the prevalence drops, the posterior probability becomes even smaller because false positives increasingly dominate. This highlights the impact of disease prevalence on test reliability.

### Comment on your results.

As the prevalence drops, the posterior probability becomes even smaller because false positives increasingly dominate. This highlights the impact of disease prevalence on test reliability.At a prevalence of 2 in 1,000, there is still a moderately higher chance of infection (posterior probability of 0.0208) compared to 15 in 100,000 (posterior probability of 0.0016). 


## QUESTION 2
Assume the following losses for Marjorie:
• Visit relative, not infected loss = 0
• Visit relative, infected loss = 100
• Do not visit relative, not infected loss = 1
• Do not visit relative, infected loss = 5

Suppose Marjorie is deciding whether to visit her relative and if so whether to test for the disease 
before visiting. 

### If the prior probability that Marjorie has the disease is 200 in 100,000, find the policy that minimizes expected loss. That is, given each of the possible test results, should Marjorie visit her relative?

```{r}
# Update prevalence
prevalence_3 <- 200 / 100000     
P_D1 <- prevalence_3             
P_D0 <- 1 - prevalence_3         

# Compute posterior probabilities for positive and negative tests
P_T1 <- (P_D1_given_T1 * P_D1) + (P_D0_given_T1 * P_D0)
P_infected_given_positive <- (P_D1_given_T1 * P_D1) / P_T1
P_not_infected_given_positive <- (P_D0_given_T1 * P_D0) / P_T1

P_T0 <- (P_D1_given_T0 * P_D1) + (P_D0_given_T0 * P_D0)
P_infected_given_negative <- (P_D1_given_T0 * P_D1) / P_T0
P_not_infected_given_negative <- (P_D0_given_T0 * P_D0) / P_T0

# Define losses
Visit_D0_loss <- 0              
Visit_D1_loss <- 100           
No_Visit_D0_loss <- 1           
No_Visit_D1_loss <- 5           

### Without Testing
Loss_Visit_No_Test <- (P_D0 * Visit_D0_loss) + (P_D1 * Visit_D1_loss)
Loss_No_Visit_No_Test <- (P_D0 * No_Visit_D0_loss) + (P_D1 * No_Visit_D1_loss)

# Print results for no testing
cat("\nExpected loss without testing:")
cat("\n- Visit relative:", Loss_Visit_No_Test)
cat("\n- Do not visit relative:", Loss_No_Visit_No_Test, "\n")

### With Testing
Loss_Visit_Positive <- (P_infected_given_positive * Visit_D1_loss) + (P_not_infected_given_positive * Visit_D0_loss)
Loss_No_Visit_Positive <- (P_infected_given_positive * No_Visit_D1_loss) + (P_not_infected_given_positive * No_Visit_D0_loss)

Loss_Visit_Negative <- (P_not_infected_given_negative * Visit_D0_loss) + (P_infected_given_negative * Visit_D1_loss)
Loss_No_Visit_Negative <- (P_not_infected_given_negative * No_Visit_D0_loss) + (P_infected_given_negative * No_Visit_D1_loss)

# Print results for testing
cat("\nExpected loss with testing:")
cat("\n- Visit relative, positive test:", Loss_Visit_Positive)
cat("\n- Do not visit relative, positive test:", Loss_No_Visit_Positive)
cat("\n- Visit relative, negative test:", Loss_Visit_Negative)
cat("\n- Do not visit relative, negative test:", Loss_No_Visit_Negative, "\n")
```
Without testing, the decision to visit results in a lower expected loss because the probability of infection is low , and the loss associated with visiting while infected is offset by the small probability of infection.

Testing provides additional information, allowing Marjorie to reduce her expected loss by acting based on the test result. For instance:
    - A negative test result suggests it is safer to visit, leading to a minimal expected loss (0.0327).

    - A positive test result suggests not visiting, which still results in a significant expected loss due to the possibility of false positives.

### Find the EVSI for 200 in 100,000
```{r}
# Conditional Expected Losses
EL_no_test <- min(Loss_Visit_No_Test , Loss_No_Visit_No_Test)
EL_with_positive <- min(Loss_Visit_Positive, Loss_No_Visit_Positive)
EL_with_negative <- min(Loss_Visit_Negative, Loss_No_Visit_Negative)

# Expected Loss With Testing
EL_with_test <- (P_T1 * EL_with_positive) + (P_T0 * EL_with_negative)

# Expected Value of Sample Information (EVSI)
EVSI <- EL_no_test - EL_with_test

# Output
cat("Expected Loss Without Testing:", EL_no_test, "\n")
cat("Expected Loss With Testing:", EL_with_test, "\n")
cat("Expected Value of Sample Information (EVSI):", EVSI, "\n")
```
The EVSI quantifies the value of testing. In this scenario, testing reduces expected loss from 0.2 (without testing) to 0.1183 (with testing), yielding an EVSI of 0.0817.

### Repeat for a prior probability of 15 in 100,000. Discuss.
```{r}
# Update prevalence for 15 in 100,000
prevalence_15 <- 15 / 100000     
P_D1 <- prevalence_15           
P_D0 <- 1 - prevalence_15       

# Compute posterior probabilities for positive and negative tests
P_T1 <- (P_D1_given_T1 * P_D1) + (P_D0_given_T1 * P_D0)
P_infected_given_positive <- (P_D1_given_T1 * P_D1) / P_T1
P_not_infected_given_positive <- (P_D0_given_T1 * P_D0) / P_T1

P_T0 <- (P_D1_given_T0 * P_D1) + (P_D0_given_T0 * P_D0)
P_infected_given_negative <- (P_D1_given_T0 * P_D1) / P_T0
P_not_infected_given_negative <- (P_D0_given_T0 * P_D0) / P_T0

# Define losses
Visit_D0_loss <- 0              
Visit_D1_loss <- 100           
No_Visit_D0_loss <- 1           
No_Visit_D1_loss <- 5           

### Without Testing
Loss_Visit_No_Test <- (P_D0 * Visit_D0_loss) + (P_D1 * Visit_D1_loss)
Loss_No_Visit_No_Test <- (P_D0 * No_Visit_D0_loss) + (P_D1 * No_Visit_D1_loss)

# Print results for no testing
cat("\nExpected loss without testing (15 in 100,000):")
cat("\n- Visit relative:", Loss_Visit_No_Test)
cat("\n- Do not visit relative:", Loss_No_Visit_No_Test, "\n")

### With Testing
Loss_Visit_Positive <- (P_infected_given_positive * Visit_D1_loss) + (P_not_infected_given_positive * Visit_D0_loss)
Loss_No_Visit_Positive <- (P_infected_given_positive * No_Visit_D1_loss) + (P_not_infected_given_positive * No_Visit_D0_loss)

Loss_Visit_Negative <- (P_not_infected_given_negative * Visit_D0_loss) + (P_infected_given_negative * Visit_D1_loss)
Loss_No_Visit_Negative <- (P_not_infected_given_negative * No_Visit_D0_loss) + (P_infected_given_negative * No_Visit_D1_loss)

# Print results for testing
cat("\nExpected loss with testing (15 in 100,000):")
cat("\n- Visit relative, positive test:", Loss_Visit_Positive)
cat("\n- Do not visit relative, positive test:", Loss_No_Visit_Positive)
cat("\n- Visit relative, negative test:", Loss_Visit_Negative)
cat("\n- Do not visit relative, negative test:", Loss_No_Visit_Negative, "\n")

# Conditional Expected Losses
EL_no_test <- min(Loss_Visit_No_Test, Loss_No_Visit_No_Test)
EL_with_positive <- min(Loss_Visit_Positive, Loss_No_Visit_Positive)
EL_with_negative <- min(Loss_Visit_Negative, Loss_No_Visit_Negative)

```
Similar to the case when the infection prevalence is 200 in 100,000, without testing, the decision to visit results in a lower expected loss. This is because the probability of infection is extremely low, and the loss associated with visiting while infected is offset by the small probability of infection.

With testing which provides an even better decision we conclude the following :-
    - Negative test: Visiting has a very low expected loss (0.0024), confirming it's safe.

    - Positive test: Visiting has a higher loss (0.159), reflecting the increased risk, while not visiting avoids that risk.

Based on the analysis, the best decision is to visit the relative if the test result is negative, as it results in the lowest expected loss (0.0024).
    
### Find the EVSI for 15 in 100,000
```{r}
# Conditional Expected Losses
EL_no_test <- min(Loss_Visit_No_Test, Loss_No_Visit_No_Test)
EL_with_positive <- min(Loss_Visit_Positive, Loss_No_Visit_Positive)
EL_with_negative <- min(Loss_Visit_Negative, Loss_No_Visit_Negative)

# Expected Loss With Testing
EL_with_test <- (P_T1 * EL_with_positive) + (P_T0 * EL_with_negative)

# Expected Value of Sample Information (EVSI)
EVSI <- EL_no_test - EL_with_test

# Output results
cat("\nExpected Loss Without Testing (15 in 100,000):", EL_no_test, "\n")
cat("Expected Loss With Testing (15 in 100,000):", EL_with_test, "\n")
cat("Expected Value of Sample Information (EVSI, 15 in 100,000):", EVSI, "\n")
```
At a very low prevalence (15 in 100,000), the expected losses with and without testing are identical. Testing provides little additional value in this setting because the disease's rarity makes the probability of infection so small that the decision (visit or not visit) is almost unaffected by the test result.

## QUESTION 3

### For the decision of whether Marjorie should visit her relative, find the range of prior probabilities for which taking the at-home test results in lower expected loss than ignoring or not taking the test (assuming the test is free).Discuss your results

Testing provides value when the EVSI > 0, which happens when the prior probability is high enough to make the test results informative.
For very low prior probabilities (like 15 in 100,000), the test offers no significant additional information.
For moderate to higher prior probabilities (like 200 in 100,000), testing reduces expected loss by refining the decision.
